Combining living and fossil taxa into phylogenies: the missing data issue

### 2014-05-14 Manuscript/TEM_draft.tex
Paper draft in proper latex

%Runing head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: Missing data in Total Evidence Supermatrices %RH = Running Head

@@

%Title
\noindent{\Large \bf Effect of missing data in supermatrices containing living and fossil taxa on topological accuracy}
\bigskip

@@

%Abstract
\subsubsection{Abstract}
Living species represent less than 1\% of all species that have ever lived. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology. This fact has led to an increasing consensus among scientists that both fossil and living taxa must be included in macroevolutionary studies. One approach, the Total Evidence Method, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both fossil and living taxa at the tips. Although the Total Evidence Method seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

In this study we assess the effect of missing data on tree topologies inferred from total evidence supermatrices. Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the supermatrix: (1) the proportion of living taxa with no morphological data, (2) the amount of missing data in the fossil record and (3) the overall number of morphological characters in the supermatrix. We find that, in a Bayesian framework, difficulties in recovering a stable topology are mainly driven by the missing data in the molecular part of the matrix (for which fossil taxa have no data). In a Maximum Likelihood framework, however, topology is not directly affected by missing data \textit{per se}, but by the number of morphological characters shared among the taxa. Therefore, the two main drivers of incorrect topologies are the overall number of morphological characters and the number of living species with no morphological data.

Our results suggest that, in order to use total evidence methods, one should reduce the missing data in the morphological part of the supermatrix for living species and use a Maximum Likelihood framework to fix the topology prior to the overall Bayesian phylogenetic inference process. We apply our method to a comprehensive data set of both living and fossil primates. We find that using this integrative method modifies previous estimates of rates of body mass evolution within primates.

\noindent (Keywords: bla, bla, bla)\\

@@

%We don't use a heading for the introduction. The nature of this section is implied by its position following the abstract and preceding the first major section heading.
Living species represent less than 1\% of all species that have ever lived \citep{novacek1992ext,raup1993extinction}. However, the majority of macroevolutionary studies focus solely on living species \citep{cooperwhat2009,meredithimpacts2011}.
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as species richness gradients, e.g. extant clade diversity may ignore past diversification patterns \citep{shoshani1996proboscidea}; biogeographical history, e.g. extant biogeographical patterns are improved when integrating fossil data \citep{metcalfintegrating2014}; or paleoecology, e.g. extant clades niches might have differed greatly through time \citep{youngthe2010}. %Rewrite that part
These factors have led to increasing consensus among scientists that fossil taxa must be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}. However, to do this we need to be able to place living and fossil taxa into the same phylogenies which still remains complex.

Three main approaches have been used for combining fossil and living taxa data in phylogenies. These approaches differ in whether they treat fossil taxa as nodes or tips and how much of the available data is used (i.e. age only or age and morphology). Classical cladistic methods use morphological data from both fossil and living taxa and treat each taxon as a tip \citep{simpson1945}. This approach is commonly used by paleontologists but it ignores
the majority of %You proposed to remove that. The point here was to not get in trouble with modern paleontologist that feel "clever" using DNA as a binary character like "Presence/absence of this SINE/that microsatellite"…
the additional molecular data available from living species. Neontologists, on the other hand, more commonly use only molecular data from living species to build phylogenies. Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these analyses and only their occurrence age is used to time calibrate phylogenies \citep{zuckerkandl1965}. There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} %Should I develop this improvements here? In the end this improvements are more on the inference methods than on the type of data used which is more the point here.
as well as much debate about the "best" approach to use \citep[e.g.][]{spencerefficacy2013}. A final class of methods, known as total evidence methods use molecular data from living taxa and morphological data from both living and fossil taxa, and treats every taxon as a tip on the phylogeny \citep{eernissetaxonomic1993}.Here we focus on these total evidence methods because they have been recently
successfully developed and applied to empirical data %needs a better justification
\citep{pyrondivergence2011,ronquista2012,schragocombining2013}. Although the total evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data. In particular it requires morphological data form both living and fossil taxa, both of which are scarce. Therefore total evidence approaches are likely to suffer from having lots of missing data which may affect their ability to infer correct phyogenies. % Trevor: Missing data also influences support values that trend to decrease

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013}. Missing molecular data has been seen by some authors as an issue because it can decrease the phylogenetic signal in some parts of the tree, especially when using large supermatrices \citep{lemmonthe2009}. However other authors do not see missing molecular data as a major issue because the phylogenetic signal is more likely to increase by having at least a 
"modest" % explain
number of highly covered genes (~50\% - \citet{rouresite-specific2011}), a higher number of taxa (especially slowly evolving taxa or taxa close to the outgroup) and by choosing more adequate models of sequence evolution rather than by reducing the amount of missing data \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as major or minor issue for accurately infering phylogenies \citep{wiensmissing2003,sansomfossilization2013}. Because soft-tissues characters are rarely preserved in the fossil record, missing data is found in soft tissues, i.e. it is not randomly distributed, which can lead to biased placement of fossil taxa in phylogenies \citep{sansomfossilization2013}. However, the phylogenetic signal is not related to the level of missing data per se but to the number of informative characters per taxa, therefore missing data is less an issue than the number of shared informative characters \citep{wiensmissing2003}.
Although not a major problem separately \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,rouresite-specific2011}, missing data in molecular and morphological supermatrices may become an issue when combined for example in a total evidence type supermatrices and no attempt has been made to study the impact of this issue until now. %Does not flow with the rest

Here we assess the effect of missing data on tree topology infered from total evidence supermatrices. The molecular part of a total evidence supermatrix contains no fossil taxa so will act like a classical molecular supermatrix \citep{ronquista2012}. The effect of missing data on such matrices is well known, therefore, we only focus on the missing data issue in the morphological part of the supermatrix. Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the supermatrix: (1) the proportion of living taxa with no morphological data, (2) the amount of missing data in the fossil taxa (i.e. the preservation quality of the fossil record) and (3) the overall number of morphological characters for both living and fossil taxa in the supermatrix. We assess how changing the values of these three parameters affects the topology of total evidence method phylogenies.

@@

\section{Methods} %no material?
To explore the effect of missing data on total evidence trees topologies we used the following protocol (note that we explain each step in detail below this general outline -Fig. ~\ref{Fig_Outline}):
\begin{enumerate}
\item{Generating the matrix}
\label{step:generate_matrix}
We built a randomly generated birth death tree to infer a "complete" matrix containing both molecular and morphological data for living and fossil taxa.
\item{Removing data}
\label{step:remove_data}
We removed data from the "complete" matrix to simulate the effects of missing data by modifying three parameters (1) the proportion of missing living taxa ($M_{L}$), (2) the proportion of missing data in the fossil taxa ($M_{F}$) and (3) the proportion of missing morphological characters ($M_{C}$).
\item{Building phylogenies}
\label{step:build_phylo}
We infered Bayesian phylogenetic trees from the "complete" matrix and from the matrices containing missing data.
\item{Comparing topologies}
\label{step:compare_topo}
We then compared the trees obtained from the matrices containing missing data to the ones obtained from the "complete" matrix to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on the topologies of the phylogenies we estimated.
\end{enumerate}
To measure the effect of missing data’s distribution, we repeated steps ~\ref{step:generate_matrix} to ~\ref{step:compare_topo} with the exact same fixed parameters 50 times.

@@

\subsection{Removing data}
Once we obtained the "complete" matrix we modified it to get a set of matrices with missing data. We randomly replaced data with “?” in the morphological part of the matrices according to the following parameters (Fig. ~\ref{Fig_RemoveData}):

@@

In practice, each parameter represent a way of removing data in the morphological part of the matrix: $M_{L}$ removes a proportion of rows from the living taxa; $M_{F}$ removes a proportion of cells from the fossil taxa; and $M_{C}$ removes a proportion of columns across both living and fossil taxa. Note that $M_{L}$ is different to $M_{F}$ not only because of the region of the supermatrix affected: for $M_{L}$, all the morphological data of a proportion of the living taxa is removed (i.e. removing rows), as for $M_{F}$, a proportion of data is removed across the whole of the morphological matrix for fossil taxa (i.e. removing cells).

@@

\subsection{Comparing topologies}
We compared the topology of the "missing-data" trees infered from the matrices with missing data to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$. Note that we only investigate differences in topology and not in branch length because the aim of this study is to look at the effect of missing data on the topology of trees infered from supermatrices with living and fossil species and molecular and morphological characters. To compare the topology of the resulting trees, we used two metrics to assess number of conserved taxa and clades position using respectively the Triple \citep{dobson1975triplets} and the Robison-Fould \citep{RF1981} distance. We normalised the two metrics using the Normalised Tree Similarity index \citep{Bogdanowicz2012} to generalise our results for any n number of taxa. The two metrics and the index are detailed below.

\subsubsection{Triple distance ($T_{x,y}$) \citep{dobson1975triplets}}
This metric measures the number of different subtrees of three clades between two given trees. Each triplet can be written as $I_{ijk}$=(\textit{ijk}). Where $I_{ijk}$ is equal to 0 if the the two triplets (\textit{ijk}) are the same in the two trees otherwise $I_{ijk}$ is equal to 1. For any rooted binary tree there are only three possible combinations per triplets: ((\textit{j},\textit{k}),\textit{i});, ((\textit{i},\textit{k}),\textit{j}); and ((\textit{i},\textit{j}),\textit{k}); \citep{johnson1998}. If the trees used are not fully binary, a fourth triplet combination is possible: (\textit{i},\textit{j},\textit{k});.One can calculate $S_n$, the triplet distance between two trees as:
\begin{equation}
S_n = \sum_{ijk} I_{ijk}
\end{equation}
Where:
\begin{equation}
\sum_{ijk} = \binom{n}{4} = \frac{n!}{4!(n-4)!}
\end{equation}
And where \textit{n} is the number of taxa in both trees (modified from \citet{critchlowthe1996}). If $S_n$=0, the trees are the same (i.e. no taxa as been displaced). When $S_n$ = $\binom{n}{4}$, the trees are the most different possible (i.e. every taxa as been displaced). This metric therefore illustrates the amount of displaxed taxa but is less sensitive to the placement of individual taxa and to taxa of highly uncertain placement (e.g. fossil taxa) than the Robinson-Foulds distance \citep{critchlowthe1996,johnson1998,wiensmissing2003} and can therefore be used as a proxy to estimate the robustness of the tree to flying taxa (see supplementaries). %link missing

@@

\subsection{Empirical data}
We also compared the results obtained from simulated data by using \citet{ronquista2012} empirical data. The supermatrix contains 67 living species plus one outgroup and 45 fossil species of Hymenopteras with 5097 molecular characters and 354 morphological characters. From the 68 living species used in the supermatrix, only 66 had molecular data %missing???
, we therefore treated these 66 taxa as "living" taxa and all the other 47 as "fossil" taxa. We treated the matrix in the exact same way as described in step 2 and 3 resulting in 125 supermatrices with various amount of missing data and the same number of Maximum Likelihood and Bayesian trees. We used the same settings as for the simulated data in the Maximum Likelihood framework. For the Bayesian inferences however, we didn’t used any priors except that we provided a starting tree with the topology of the 68 living species (topology with the highest posterior probability from non-clock analysis - \citet{ronquista2012}). Contrary to \citet{ronquista2012} analysis, we didn’t performed any clock analysis since we were only interested in the topology of the infered tree and not the branch length.

\section{Results}
%Baysian inference doesn’t recover the topology, whatever the amount of data
%Bayesian inference with or without the fossil/living doesn’t recover the topology, whatever the amount of data
%Likelihood inference recovers the topology depending on the amount of data
%Likelihood with only living recovers the topology really well
%Likelihood with only fossil??

@@

>>>>

%Runing head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: Missing data in Total Evidence matrices %RH = Running Head

@@

%Title
\noindent{\Large \bf Effect of missing data in matrices containing living and fossil taxa on topological accuracy}
\bigskip

@@

%IMPORTANT:
%Note on display in the text file: sentences or sub-parts of sentences (e.g. enumerations) are delimited by a newline for clarity.
%However, true new lines (i.e. in the pdf) are delimited by one or more empty lines.

%Abstract
\subsubsection{Abstract}
Living species represent less than 1\% of all species that have ever lived. Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as trends in species richness, biogeographical history or paleoecology.
This fact has led to an increasing consensus among scientists that both fossil and living taxa must be included in macroevolutionary studies.
One approach, the total evidence approach, uses molecular data from living taxa and morphological data from both living and fossil taxa to infer phylogenies with both fossil and living taxa at the tips.
Although the total evidence approach seems very promising, it requires a lot of data and is therefore likely to suffer from missing data issues which may affect its ability to infer correct phylogenies.

In this study we assess the effect of missing data on tree topologies inferred from total evidence matrices.
Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the matrix:
(1) the proportion of living taxa with no morphological data,
(2) the amount of missing data in the fossil record and
(3) the overall number of morphological characters in the matrix.
We find that, in a Bayesian framework, difficulties in recovering a stable topology are mainly driven by the missing data in the molecular part of the matrix (for which fossil taxa have no data).
In a Maximum Likelihood framework, however, topology is not directly affected by missing data \textit{per se}, but by the number of morphological characters shared among the taxa.
Therefore, the two main drivers of incorrect topologies are the overall number of morphological characters and the number of living species with no morphological data.

Our results suggest that, in order to use total evidence approaches, one should reduce the missing data in the morphological part of the matrix for living species and use a Maximum Likelihood framework to fix the topology prior to the overall Bayesian phylogenetic inference process.
We apply our method to a comprehensive data set of both living and fossil primates. %Not at all!
We find that using this integrative method modifies previous estimates of rates of body mass evolution within primates.

\noindent (Keywords: missing data, total evidence, bla)\\

@@

%Syst.Biol: We don't use a heading for the introduction. The nature of this section is implied by its position following the abstract and preceding the first major section heading.
Living species represent less than 1\% of all species that have ever lived \citep{novacek1992ext,raup1993extinction}.
However, the majority of macroevolutionary studies focus solely on living species \citep{cooperwhat2009,meredithimpacts2011,healyecology2014}. %or is citing of your own papers too douchebagy in the second sentence?
Ignoring fossil taxa may lead to misinterpretation of macroevolutionary patterns and processes such as timing of diversification events \citep[e.g.][]{pyrondivergence2011}, relationships among lineages \citep[e.g.][]{manosphylogeny2007} or niche occupancy \citep[e.g.][]{pearmanniche2008}.
These factors have led to increasing consensus among scientists that fossil taxa must be included in macroevolutionary studies \citep{jacksonwhat2006,quentaldiversity2010,dietlconservation2011,slaterunifying2013,fritzdiversity2013}.
However, to do this we need to be able to place living and fossil taxa into the same phylogenies which still remains complex despite an increasing number of significant efforts \citep{pyrondivergence2011,ronquista2012,schragocombining2013}. %Not a nice sentence

Three main approaches have been used for combining fossil and living taxa data in phylogenies.
These approaches differ mainly in whether they treat fossil taxa as tips or as nodes in the phylogeny and which part of the available data of the fossil is used (i.e. the fossil occurrence age only or both the age and the morphology).
Classical cladistic methods uses matrices containing morphological data from both fossil and living taxa and allows to treat each taxon as a tip in the phylogeny; the relation between the taxa can be infered using optimal criteria such as the maximum parsimony criterion \citep{simpson1945}.
This approach is commonly used by paleontologists but it ignores the additional molecular data available from living species.
Neontologists, on the other hand, more commonly use probabilistic methods (e.g. Maximum Likelihood or Bayesian) based on matrices containing only molecular data from living species to infer phylogenies.
Because fossil taxa do not usually have available DNA, fossils are used as nodes rather than tips in these analyses and only their occurrence age can be used to time calibrate phylogenies \citep{zuckerkandl1965}.
There have been great improvements in the theory and application of these two approaches \citep[e.g.][]{bapsta2013,stadlerdating2013,heaththe2013} as well as much debate about the "best" approach to use \citep[e.g.][]{spencerefficacy2013} but none of them uses all the available data.

A final approach, known as total evidence approach use matrices containing molecular data from living taxa and morphological data from both living and fossil taxa. %There might be a philosophico-semantic distinction to do here. Is TEM a method or an approach? I'll say that it is an approach because it's more like an idea (put every thing together) than a way to do (how to put everything together). It makes more sense to me compared to a real method e.g. a probabilistic method such as ML where the method states precisely how to do stuff.
It allows to use probabilistic methods, to treat every taxon as a tip on the phylogeny and to use the occurrence age of the fossil to time calibrate the phylogeny \citep{eernissetaxonomic1993}.
Here we focus on this total evidence approach because they have been recently successfully developed and applied to empirical data \citep{pyrondivergence2011,ronquista2012,schragocombining2013}.  %needs a better justification
Although the total evidence approach seems very promising, there is one big drawback in using this approach: it requires a lot of data.
In particular it requires morphological data form both living and fossil taxa, both of which are scarce.
Therefore total evidence approaches are likely to suffer from having lots of missing data which may affect their ability to infer correct phylogenies. % Trevor: Missing data also influences support values that trend to decrease

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013}.
Missing molecular data has been seen by some authors as an issue because it can decrease the phylogenetic signal in some parts of the tree, especially when using large matrices \citep{lemmonthe2009}.
However other authors do not see missing molecular data as a major issue because the phylogenetic signal is more likely to increase by having at least a "modest" number of highly covered genes (i.e. approximatively half of the genes - \citet{rouresite-specific2011}), a higher number of taxa (especially slowly evolving taxa or taxa close to the outgroup) and by choosing more adequate models of sequence evolution rather than by reducing the amount of missing data \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}.
Similarly, depending on the study, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study \citep{wiensmissing2003,sansomfossilization2013}.
Because soft-tissues characters are rarely preserved in the fossil record, missing data is mainly found in soft tissues characters, and is therefore not randomly distributed, which can lead to biased placement of fossil taxa in phylogenies \citep{sansomfossilization2013}.
However, the phylogenetic signal is not related to the level of missing data \textit{per se} but to the number of informative characters per taxa, therefore missing data is less an issue than the number of shared informative characters \citep{wiensmissing2003}.
Although missing data as been shown to be no major problem separately in molecular and morphological matrices \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,rouresite-specific2011} it may become an issue when combined in a total evidence type matrix and no attempt has been made to study the impact of this issue on phylogenetic inference until now (especially because a major part of the matrix is missing - the molecular data from the fossil taxa). %Does not flow with the rest

Here we assess the effect of missing data on tree topology inferred from total evidence matrices.
The molecular part of a total evidence matrix contains no fossil taxa so will act like a classical molecular matrix \citep{ronquista2012}.
The effect of missing data on such matrices is well known, therefore, we only focus on the missing data issue in the morphological part of the matrix.
Using simulations we investigate three major factors that directly affect the completeness of the morphological part of the matrix:
(1) the proportion of living taxa with no morphological data,
(2) the amount of missing data in the fossil taxa (i.e. the preservation quality of the fossil record) and
(3) the overall number of morphological characters for both living and fossil taxa in the matrix.
We assess how changing the values of these three parameters affects the topology of total evidence approach phylogenies.
%This paragraph is crap

@@

\section{Methods} %no material needed?
To explore the effect of missing data on total evidence trees topologies we used the following protocol (note that we explain each step in detail below this general outline -Fig. ~\ref{Fig_Outline}):
\begin{enumerate}
\item{Generating the matrix}
\label{step:generate_matrix}
We built a randomly generated birth death tree to infer a "complete" matrix containing both molecular and morphological data for living and fossil taxa.
\item{Removing data}
\label{step:remove_data}
We removed data from the "complete" matrix to simulate the effects of missing data by modifying three parameters (1) the proportion of missing living taxa ($M_{L}$), (2) the proportion of missing data in the fossil taxa ($M_{F}$) and (3) the proportion of missing morphological characters ($M_{C}$).
\item{Building phylogenies}
\label{step:build_phylo}
We infered Bayesian phylogenetic trees from the "complete" matrix and from the matrices containing missing data.
\item{Comparing topologies}
\label{step:compare_topo}
We then compared the trees obtained from the matrices containing missing data to the ones obtained from the "complete" matrix to assess the influence of each parameter ($M_{L}$, $M_{F}$, $M_{C}$ and their interactions) on the topologies of the phylogenies we estimated.
\end{enumerate}
To measure the effect of missing data’s distribution, we repeated steps ~\ref{step:generate_matrix} to ~\ref{step:compare_topo} with the exact same fixed parameters 50 times.

@@

\subsection{Removing data}
Once we obtained the "complete" matrix we modified it to get a set of matrices with missing data
 We randomly replaced data with "?" in the morphological part of the matrices according to the following parameters (Fig. ~\ref{Fig_RemoveData}):

@@

In practice, each parameter represent a way of removing data in the morphological part of the matrix: $M_{L}$ removes a proportion of rows from the living taxa; $M_{F}$ removes a proportion of cells from the fossil taxa; and $M_{C}$ removes a proportion of columns across both living and fossil taxa.
Note that $M_{L}$ is different to $M_{F}$ not only because of the region of the matrix affected: for $M_{L}$, all the morphological data of a proportion of the living taxa is removed (i.e. removing rows), as for $M_{F}$, a proportion of data is removed across the whole of the morphological matrix for fossil taxa (i.e. removing cells).

@@

\subsection{Comparing topologies}
We compared the topology of the "missing-data" trees infered from the matrices with missing data to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$.
Note that we only investigate differences in topology and not in branch length because the aim of this study is to look at the effect of missing data on the topology of trees infered from matrices with living and fossil species and molecular and morphological characters.
To compare the topology of the resulting trees, we used two metrics to assess number of conserved taxa and clades position using respectively the Triple \citep{dobson1975triplets} and the Robison-Fould \citep{RF1981} distance.
We normalised the two metrics using the Normalised Tree Similarity index \citep{Bogdanowicz2012} to generalise our results for any n number of taxa. The two metrics and the index are detailed below.

\subsubsection{Triple distance ($T_{x,y}$) \citep{dobson1975triplets}}
This metric measures the number of different subtrees of three clades between two given trees.
Each triplet can be written as $I_{ijk}$=(\textit{ijk}).
Where $I_{ijk}$ is equal to 0 if the the two triplets (\textit{ijk}) are the same in the two trees otherwise $I_{ijk}$ is equal to 1.
For any rooted binary tree there are only three possible combinations per triplets: ((\textit{j},\textit{k}),\textit{i});, ((\textit{i},\textit{k}),\textit{j}); and ((\textit{i},\textit{j}),\textit{k}); \citep{johnson1998}.
If the trees used are not fully binary, a fourth triplet combination is possible: (\textit{i},\textit{j},\textit{k});.
One can calculate $S_n$, the triplet distance between two trees as:
\begin{equation}
S_n = \sum_{ijk} I_{ijk}
\end{equation}
Where:
\begin{equation}
\sum_{ijk} = \binom{n}{4} = \frac{n!}{4!(n-4)!}
\end{equation}
And where \textit{n} is the number of taxa in both trees (modified from \citet{critchlowthe1996}).
If $S_n$=0, the trees are the same (i.e. no taxa as been displaced). When $S_n$ = $\binom{n}{4}$, the trees are the most different possible (i.e. every taxa as been displaced).
This metric therefore illustrates the amount of displaxed taxa but is less sensitive to the placement of individual taxa and to taxa of highly uncertain placement (e.g. fossil taxa) than the Robinson-Foulds distance \citep{critchlowthe1996,johnson1998,wiensmissing2003} and can therefore be used as a proxy to estimate the robustness of the tree to flying taxa (see supplementaries). %link missing

@@

\subsection{Inference testing}
How we tested the inference (vertical/horizontal/corner).

\subsection{Empirical data}
We also compared the results obtained from simulated data by using \citet{ronquista2012} empirical data.
The matrix contains 67 living species plus one outgroup and 45 fossil species of Hymenopteras with 5097 molecular characters and 354 morphological characters.
From the 68 living species used in the matrix, only 66 had molecular data, we therefore treated these 66 taxa as "living" taxa and all the other 47 as "fossil" taxa.
We treated the matrix in the exact same way as described in step 2 and 3 resulting in 125 matrices with various amount of missing data and the same number of Maximum Likelihood and Bayesian trees.
We used the same settings as for the simulated data in the Maximum Likelihood framework.
For the Bayesian inferences however, we didn’t used any priors except that we provided a starting tree with the topology of the 68 living species (topology with the highest posterior probability from non-clock analysis - \citet{ronquista2012}).
Contrary to \citet{ronquista2012} analysis, we didn’t performed any clock analysis since we were only interested in the topology of the infered tree and not the branch length.

\section{Results}
The ability of recovering the best topology in maximum likelihood method is function of the amount of data missing.
However, in Bayesian, topology is badly (if not at all) recovered independently of the amount of data missing.

@@

What about the "best" trees with fairly bad bootstrap support values in ML?
This is probably due to the low number of characters used in the simulation.
The number of morphological characters ($10^2$) is in the same order of magnitudes of empirical studies (cite) but decreases unrealistically with the $M_C$ parameter. %Extract the mean number of characters per matrices from the 100 random ones
The number of molecular characters ($10^3$) is at least two orders of magnitude lower than in empirical studies that can reach $10^6$ characters (cite).

Two different matrices from the "True" tree gives two different phylogenetic interpretations (i.e. seqGen and ape performs badly at generating a data matrix from a tree)?
That seems to not really be a problem since we are interested in recovering topology with missing data, not assessing if phylogenetic matrices generation software performs correctly.
Also, the "noise" that might be created by this potential "bad" phylogenetic matrices generation software is probably diminished by the number of replicates (i.e. 50 chains).

Different phylogenetic signal between the morphological and the molecular matrix?
That is not really a problem since the phenotypic evolution is not necessary correlated to the genotypic one.
Also it is important to note that both information are not exactly the same.
The main difference between molecular and morphological characters is that, because of historical and practical reasons, a morphological character matrix do no contains autapomorphies and invariant characters which are usually more common than synampomorphies in molecular character matrices (cite).

What about homoplasy in the morphological matrix?
Here, we made the assumption that theoretically, morphological characters are randomly distributed on an organism however it seems clear that empirical morphological data does not act randomly
If one states that morphological characters accumulate through time in the same way as the majority of the molecular characters (neutral theory, cite) then, homoplasic characters are expected to appear randomly through time.
Therefore, homoplasy is expected to be more important (by chance) in bigger morphological matrices \citep{davalosintegrating2014}.

\section{Conclusion}
A Bayesian approach fails to recover accurate topology in a total evidence approach framework, whatever the amount of missing data.
We think this failure is due to the intrinsic structure of the Total Evidence matrices that includes a vast amount of not randomly distributed missing data (i.e. the molecular part of the matrix for the fossil taxa).
This missing data is equal to the number of fossil taxa $\times$ the number of molecular characters leading to a vast amount of likely trees to be sampled in the Bayesian mcmc tree building.
However, one can use a Maximum Likelihood approach to obtain the likeliest topology and to use this one as a fixed.

\section{Acknowledgements}
We would like to thank Trevor Hodkinson %Also Gavin Thomas, Emmanuel Douzery and Frédéric Delsuc if not authors.
for his useful comments on the simulation protocol and Paddy Doyle for the assistance on using the computer cluster.
All calculations were performed on the Lonsdale cluster maintained by the Trinity Centre for High Performance Computing.
This cluster was funded through grants from Science Foundation Ireland.
